Ultra-compact, widely bandwidth-tunable filter using cascaded silicon photonic contra-directional couplers

### 2015-05-21 CacadedTunableFilter.tex
Update

While an unapodized (uniform) grating can be simulated with only two transfer matrices , $S_1$ and $S_2$ \cite{shi2013siliconContraDC}, an apodized one needs both of these matrices for each step along the propagation, to match the change of coupling. 
The first change is thus to modify the transfer matrix of the system to the product of all the grating elements along the propagation axis.
This calculation ideally takes shape in an infinite sum, but for computation simplicity, we use a finite number of steps with different couplings that still approximates the apodization profile with enough accuracy. 
Our simulations uses  steps of constant length $z_i-z_{i-1}=z_c$. 
The total response of a contra-DC $C(z_0,z_n)$ is calculated using the transfer-matrix method, where $E(z_0)=C(z_0,z_n) E(z_n)$.

\begin{equation}
\label{eq:CMatrix}
 C(z_0,z_n) =  \prod_{i=1}^{n}\me^{S_{1,i}(z_i-z_{i-1})}\me^{S_{2,i}(z_i-z_{i-1})}
\end{equation}

@@

On top of each contra-DC sits a metal strip acting as a microheater for thermal tuning.  
The refractive index of silicon has a temperature dependence of 1.87$\times10^{-4}$K$^{-1}$ at room temperature for wavelengths around 1550 nm~\cite{frey2006temperature}.
Without electrical input, the optical responses of the two filters are identical (neglecting fabrication uniformities) resulting in a sharp transition from the pass-band to the stop-band. 
Applying independent electrical currents on the heaters, we can shift the spectra of the contra-DCs simultaneously or differentially for wavelength or bandwidth tuning.
In the ideal case, each contra-DC should be uniformly heated so that the apodization profile does not change as the temperature varies.

@@

The device was fabricated using a CMOS-compatible technology with electron-beam lithography. 
Fiber grating couplers\cite{zhong2014focusingFGC} were used as optical IOs in the measurement. 
Fig.~\ref{fig:passive} shows the measured drop port response of the cascaded contra-DC filters. 
The measurements were normalized using the response of a pair of directly connected fiber grating couplers on the same chip.
\begin{figure}[htbp]
\centering
\includegraphics[width=.99\columnwidth]{data/Passive6}
\caption{ Spectral response of the cascaded filters without heating: 1-dB BW is 733 GHz, 3-dB BW is 788 GHz, 10-dB BW is 868 GHz, and 20-dB BW is 990 GHz. Inset shows the filter shape up to -20 dB, which is typical in communications.}
\label{fig:passive}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.99\columnwidth]{data/Band6}
\caption{Spectral response of the device for different temperatures applied to only one contra-DC: 1-dB BW tuned down to 65 GHz, 3-dB BW to 117 GHz. The temperatures are obtained from simulation fit.}
\label{fig:bandTune}
\end{figure}

By applying the same temperature variation on both contra-DCs, the center wavelength can be tuned without affecting the filter shape.
As shown in Fig.~\ref{fig:wavTune}, when the center wavelength is continually changed over 4 nm by varying the on-chip temperature, the filter shape is maintained with sharp edges. Actually, slight detuning between the cascaded contra-DCs may be used to compensate for band-edge distortions due to fabrication errors for a more symmetric filter shape. 
The power efficiency of the wavelength tuning is about 44 mW/nm, which is about twice the power consumption of the bandwidth tuning since two contra-DCs are heated simultaneously.
\begin{figure}[htbp]
\centering
\includegraphics[width=.99\columnwidth]{data/Central2}
\caption{Spectral response with the heat applied to both contra-DCs: the central wavelength is tuned from 1535 nm to 1539 nm continuously; the temperatures are calculated by comparison to simulation.}
\label{fig:wavTune}
\end{figure}

>>>>

While an unapodized (uniform) grating can be simulated with only two transfer matrices , $S_1$ and $S_2$ \cite{shi2013siliconContraDC}, an apodized one needs both of these matrices for each step along the propagation, to match the change of coupling. 
The first change is thus to modify the transfer matrix of the system to the product of all the grating elements along the propagation axis.
This calculation ideally takes shape in an infinite sum, but for computation simplicity, we use a finite number of steps with different couplings that still approximates the apodization profile with enough accuracy. 
Our simulations uses  steps of constant length $z_i-z_{i-1}=z_c$. 
%The total response of a contra-DC $C(z_0,z_n)$ is calculated using the transfer-matrix method, where $E(z_0)=C(z_0,z_n) E(z_n)$.

%\begin{equation}
%\label{eq:CMatrix}
% C(z_0,z_n) =  \prod_{i=1}^{n}\me^{S_{1,i}(z_i-z_{i-1})}\me^{S_{2,i}(z_i-z_{i-1})}
%\end{equation}

@@

On top of each contra-DC sits a metal strip acting as a microheater for thermal tuning.  
The refractive index of silicon has a temperature dependence of $\frac{\diff{n_\text{Si}}}{\diff{T}}=$ 1.87$\times10^{-4}$K$^{-1}$ at room temperature for wavelengths around 1550 nm~\cite{frey2006temperature}.
Without electrical input, the optical responses of the two filters are identical (neglecting fabrication uniformities) resulting in a sharp transition from the pass-band to the stop-band. 
Applying independent electrical currents on the heaters, we can shift the spectra of the contra-DCs simultaneously or differentially for wavelength or bandwidth tuning.
In the ideal case, each contra-DC should be uniformly heated so that the apodization profile isn't disturbed as the temperature varies.

In simulation, the temperature dependence is accounted for by increasing the effective index according to 
$\diff{n_\text{eff}}=\frac{\diff{n_\text{Si}}}{\diff{T}}*\frac{\diff{n_\text{eff}}}{\diff{n_\text{Si}}}*\diff{T}$, where $\frac{\diff{n_\text{eff}}}{\diff{n_\text{Si}}}$ is calculated using an eigenmode solver.

@@

The device was fabricated using a CMOS-compatible technology with electron-beam lithography. 
Fiber grating couplers\cite{zhong2014focusingFGC} were used as optical IOs in the measurement. 
Fig.~\ref{fig:passive} shows the measured drop port response of the cascaded contra-DC filters. 
The measurements were normalized using the response of a pair of directly connected fiber grating couplers on the same chip.
\begin{figure}[htbp]
\centering
\includegraphics[width=.99\columnwidth]{data/Passive6}
\caption{ Spectral response of the cascaded filters without heating: 1-dB BW is 733 GHz, 3-dB BW is 788 GHz, 10-dB BW is 868 GHz, and 20-dB BW is 990 GHz. Inset shows the filter shape up to -20 dB, which is the typical communication requirement.}
\label{fig:passive}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.99\columnwidth]{data/Band6}
\caption{Spectral response of the device for different temperatures applied to only one contra-DC: 1-dB BW tuned down to 65 GHz, 3-dB BW to 117 GHz. The temperatures are obtained from simulation fit. Inset shows the temperature dependence.}
\label{fig:bandTune}
\end{figure}

By applying the same temperature variation on both contra-DCs, the center wavelength can be tuned without affecting the filter shape.
As shown in Fig.~\ref{fig:wavTune}, when the center wavelength is continually changed over 4 nm by varying the on-chip temperature, the filter shape is maintained with sharp edges. Actually, slight detuning between the cascaded contra-DCs may be used to compensate for band-edge distortions due to fabrication errors for a more symmetric filter shape. 
The power efficiency of the wavelength tuning is about 44 mW/nm, which is about twice the power consumption of the bandwidth tuning since two contra-DCs are heated simultaneously.
\begin{figure}[htbp]
\centering
\includegraphics[width=.99\columnwidth]{data/Central2}
\caption{Spectral response with the heat applied to both contra-DCs: the central wavelength is tuned from 1535 nm to 1539 nm continuously; the temperatures are calculated by comparison to simulation. Inset shows the temperature dependence.}
\label{fig:wavTune}
\end{figure}
